---
title: "Missing Data and Imputation"
author: "Anthony Kamau"
toc: true
output:
  pdf_document: default
  pdf: default
---

## Introduction
This markdown tackles ad-hoc imputation methods as well as multiple imputation,
all using the `mice` package.

The following steps are performed:

   - Data processing
   - Inspection of the missingness pattern
   - Apply ad hoc imputation methods using the `mice` package
   - Apply multiple imputation using the `mice` package
   - Perform linear regression and compare the output when different imputation 
      approaches are used
   

## Load the required libraries. 
```{r setup, message=FALSE}
# Load libraries
library(tidyverse)
library(mice)
library(ggmice)
library(ISLR)

# set seed for reproducability 
set.seed(123)

# set scippen to suppress scientific notations
options(scipen = 999)
```

## Data Processing
From the ISLR package, We will use the College dataset and the following variables:

    - Out-of-state tuition
    - PhD: Pct. of faculty with Ph.D.’s
    - Terminal: Pct. of faculty with terminal degree
    - Expend: Instructional expenditure per student
    
```{r}
data_complete <- 
  College |>
  select(Outstate, PhD, Terminal, Expend)

summary(data_complete)

# This is the complete ground truth data before any missingness is introduced.
# It will be the benchmark against which every imputation method will be later 
# compared to
```
We focus on how different imputation approaches affect the linear relationship 
between `Outstate` and `Expend`. First we inspect how the relationship looks 
like on the complete dataset.

```{r}
# reg of Expend on Outstate using the full complete data
lm_complete <- lm(Expend ~ Outstate,
                  data = data_complete)

summary(lm_complete)

# Outstate coefficient (0.873), highly significant p <0.0000000000000002
# This is the gold standard benchmark since no missingness exists yet
# This is the "correct" answer every later model will be compared against 
```
## Missingness Generation (SDD/MAR mechanism)
We can now generate missingness in the variable `Expend` and `Terminal` under a 
Seen Data Dependent (SDD) mechanism, assuming that the missingness in these 
variables are dependent on the values of the `Outstate` and `PhD` respectively

```{r}
# logistic function
logistic <- function(x) (exp(x) / (1 + exp(x))) 

# number of observations
N <- nrow(data_complete) 

# predictor Outstate
misprob_out <- logistic(scale(data_complete$Outstate)) 

# predictor PhD
misprob_phd <- logistic(scale(data_complete$PhD)) 

data_missing <- 
  data_complete |> 
  mutate(
    # response indicator outstate
    R_out = 1 - rbinom(N, 1, misprob_out), 
    # response indicator PhD
    R_phd = 1 - rbinom(N, 1, misprob_phd), 
    # Missing in expend
    m_Expend = ifelse(R_out == 0, NA, Expend),
    # and in terminal
    m_Terminal = ifelse(R_phd == 0, NA, Terminal) 
  ) |> 
  # Select variables that we will use
  select(Outstate, m_Expend, m_Terminal) 


# Outstate (a variable that stays fully observed) is the driver of missingness 
# in Expend, and PhD drives missingness in Terminal
```

## Inspection of missingness
Inspect the missingness pattern using `ggmice` package function
```{r}
# missingness pattern plot
plot_pattern(data_missing)

# 219 rows are missing both variables (Expend and Terminal)
# 154 rows are only missing Expend variable
# 188 rows are only missing Terminal variable
# 216 rows have complete data on all three variables
# In total there are 780 missing entries across the dataset i.e 
# 373 missing in m_Expend (154 + 219)
# 407 missing in m_Terminal (188 + 219) 
# Out of 1,554 possible missing cells for these two variables 2.(216 + 188 + 154 + 219)
# meaning roughly half of all values in these two columns are missing. 
# Outstate itself has zero missing values (it's the fully-observed predictor of missingness).
```

## Ad Hoc Imputation Methods
### Listwise Deletion
This drops any row with a missing value and fits the regression on the 404 complete 
rows (777 - 373)
```{r}
lm_listwise <- lm(m_Expend ~ Outstate,
                  data = data_missing)

summary(lm_listwise)

# Coefficient (0.702) vs true coefficient (0.873)
# The coefficient is a bit smaller,
# Standard error (0.0415) vs true standard error (0.0345)
# the standard error somewhat larger, 
# so we introduced some bias, and reduced the power. Both could have been expected,
# due to how we generated missings and since we lost 373 cases.
# R squared (0.4155 vs true R squared (0.4526). It dropped.
```
### Regression Imputation
This replaces each missing value with its predicted value from a regression model
```{r}
imp_regression <- mice(
  data   = data_missing,
  method = "norm.predict",
  m      = 1,
  maxit  = 1
)

# Coefficient (0.703). Almost identical to listwise deletion, still biased 
# relative to the true coefficient (0.873)
# Standard error (0.0178), much smaller that the true standard error (0.0345)
# By imputing with no added noise, the method artificially shrinks the variability 
# in the data, making standard error unrealistically small. In short, it is 
# pretending to know these values more precisely than it actually does.

```

`complete()` takes an object, fills in the missing data, and returns the complete
data in a specified format.
```{r}
complete(imp_regression) |>  
  is.na() |>                
  sum()
```

```{r}
lm_regression <- with(imp_regression, lm(m_Expend ~ Outstate))

summary(lm_regression)
```

```{r}
imp_regression |> 
  ggmice(aes(x = Outstate, y = m_Expend)) +
  geom_point()

# Almost all imputed values are on a (almost) straight regression line (which 
# is how we specified the imputation model). The variation around this line
# comes from the other variables that are also used in calculating the
# predictions.
# Each imputed value is a deterministic prediction with no residual variability 
# added, all imputed points cluster tightly along the regression line rather 
# than scattering naturally like real data would.
```

### Stochastic Regression Imputation
This improves on plain regression imputation by adding random residual noise to each 
predicted value, so that imputed points scatter realistically around the regression 
line rather than sitting exactly on it.
```{r}
imp_stochastic <- mice(
  data   = data_missing,
  method = "norm.nob",
  m      = 1,
  maxit  = 1
)
```

```{r}
lm_stochastic <- with(imp_stochastic, lm(m_Expend ~ Outstate))

summary(lm_stochastic) 

# Coefficients (0.715), similar to other ad hoc methods. Still biased.
# Standard error (0.0257), bigger that regression imputation (0.0178) but still 
# smaller than the true (0.0345)
```

```{r}
imp_stochastic |> 
  ggmice(aes(x = Outstate, y = m_Expend)) +
  geom_point()

# There is considerably more variation between the imputed values as compared to
# the regression imputations. However, the current approach still treats the 
# estimated regression coeffcients as fixed, although there is also uncertainty 
# with respect to these coefficients. This uncertainty is still not taken into 
# account with stochastic regression imputation.
```

### Multiple Imputation
This creates 5 separate imputed datasets (the m=5 default), each filled in using
Predictive Mean Matching ("pmm"), iterating 5 times (maxit=5, the default) per dataset.

    - 5 imputations, method "pmm" for both m_Expend and m_Terminal
    - Predictor matrix: shows which variables are used to impute which, for example
      Outstate and m_Terminal are both used as predictors when imputing m_Expend, and vice versa
      
```{r}
imp_multiple <- mice(data_missing) 

imp_multiple

```


```{r}
# fits regression separately on all 5 imputed datasets
lm_multiple <- with(imp_multiple, lm(m_Expend ~ Outstate)) 

# The `lm_multiple` object contains the regression summaries for each of the 5
# datasets. The new object `lm_multiple` is actually of class `mira`, which 
# stands for Multiply Imputed Repeated Analyses.

```


```{r}
# combines those 5 sets of results using Rubin's pooling rules
pool_multiple <- pool(lm_multiple)  

summary(pool_multiple)

# These results show the regression coefficient and corresponding standard 
# error obtained after pooling the estimates from the five regression models
# applied on the five imputed datasets.

# Coefficient (0.853), very close to the true coefficient (0.873) by far the 
# closest of any method
# Standard error (0.0443), actually larger than the true standard error (0.0345),
# which is the expected and correct behavior. It honestly reflects the added 
# uncertainty from having real missing data, rather than artificially shrinking
# it like the ad hoc methods did.

# The complete pooled object gives the relevant pooled regression coefficients 
# and parameters, as well as the fraction of information about the coefficients
# missing due to nonresponse (`fmi`) and the proportion of the variation
# attributable to the missing data (`lambda`). The pooled fit object is of 
# class `mipo`, which stands for multiply imputed pooled object.*
```

```{r}
imp_multiple |> 
  ggmice(aes(x = Outstate, y = m_Expend)) +
  geom_point()

# imputed points blend naturally with observed points across the whole range
# Here we take all uncertainty with respect to the imputations into account 
# (i.e. uncertainty with respect to the values and with respect to the
# coefficients). 
# Moreover, we impute multiple times, so that we can assess the variability 
# between different data sets (on the level of the estimated model parameters). 
# This is often the best approach when interest is in making  inferences with 
# respect to the imputed data.
```

## Comparisson
```{r}
data.frame("method" = c("complete data", 
                        "listwise deletion", 
                        "regression imputation",
                        "stochastic regression imputation",
                        "multiple imputation"),
           "coefficient" = c(summary(lm_complete)$coefficients[2,1],
                             summary(lm_listwise)$coefficients[2,1],
                             summary(lm_regression)$estimate[2],
                             summary(lm_stochastic)$estimate[2],
                             summary(pool_multiple)$estimate[2]),
           "standard error" = c(summary(lm_complete)$coefficients[2,2],
                                summary(lm_listwise)$coefficients[2,2],
                                summary(lm_regression)$std.error[2],
                                summary(lm_stochastic)$std.error[2],
                                summary(pool_multiple)$std.error[2]))

# When comparing the results from the different approaches, it can be 
# concluded that in case of a dataset where the missingness follows the 
# mechanism Seen Data Dependent, the estimate obtained after applying multiple
# imputation is closest to the original estimate obtained from the complete
# dataset. 
#In addition, it can be concluded that the standard error obtained using multiple
# imputation is somewhat larger than the standard error obtain from the complete
# dataset. This is due to the fact that when the standard errors are pooled over
# the different imputations, not only the average is 
# taken, but the variance over the differences is also incorporated i.e. the 
# standard error also incorporates uncertainty due to the missing values. 
# In contrast, the results obtained using listwise deletion, regression imputation
# and stochastic imputation all show a biased estimate, while regression imputation
# and stochastic imputation also show a standard error that is too low.

```


^