---
title: "Missing Data and Imputation"
author: "Anthony Kamau"
toc: true
output:
  pdf_document: default
  pdf: default
---

## Introduction
This markdown tackles ad-hoc imputation methods as well as multiple imputation,
all using the `mice` package.

The following steps are performed:

   - Data processing
   - Inspection of the missingness pattern
   - Apply ad hoc imputation methods using the `mice` package
   - Apply multiple imputation using the `mice` package
   - Perform linear regression and compare the output when different imputation 
      approaches are used
   

## Load the required libraries. 
```{r setup, message=FALSE}
# Load libraries
library(tidyverse)
library(mice)
library(ggmice)
library(ISLR)

# set seed for reproducability 
set.seed(123)

# set scippen to suppress scientific notations
options(scipen = 999)
```

## Data Processing
From the ISLR package, We will use the College dataset and the following variables:

    - Out-of-state tuition
    - PhD: Pct. of faculty with Ph.D.’s
    - Terminal: Pct. of faculty with terminal degree
    - Expend: Instructional expenditure per student
    
```{r}
data_complete <- 
  College |>
  select(Outstate, PhD, Terminal, Expend)

summary(data_complete)
```
We focus on how different imputation approaches affect the linear relationship 
between `Outstate` and `Expend`. First we inspect how the relationship looks 
like on the complete dataset.

```{r}
lm_complete <- lm(Expend ~ Outstate,
                  data = data_complete)

summary(lm_complete)
```
We can now generate missingness in the variable `Expend` and `Terminal` under a 
Seen Data Dependent (SDD) mechanism, assuming that the missingness in these 
variables are dependent on the values of the `Outstate` and `PhD` respectively

```{r}
# logistic function
logistic <- function(x) (exp(x) / (1 + exp(x))) 

# number of observations
N <- nrow(data_complete) 

# predictor Outstate
misprob_out <- logistic(scale(data_complete$Outstate)) 

# predictor PhD
misprob_phd <- logistic(scale(data_complete$PhD)) 

data_missing <- 
  data_complete |> 
  mutate(
    # response indicator outstate
    R_out = 1 - rbinom(N, 1, misprob_out), 
    # response indicator PhD
    R_phd = 1 - rbinom(N, 1, misprob_phd), 
    # Missing in expend
    m_Expend = ifelse(R_out == 0, NA, Expend),
    # and in terminal
    m_Terminal = ifelse(R_phd == 0, NA, Terminal) 
  ) |> 
  # Select variables that we will use
  select(Outstate, m_Expend, m_Terminal) 
```

## Inspection of missingness
Inspect the missingness pattern using `ggmice` package function
```{r}
plot_pattern(data_missing)
```

## Ad Hoc Imputation Methods
### Listwise Deletion

```{r}
lm_listwise <- lm(m_Expend ~ Outstate,
                  data = data_missing)

summary(lm_listwise)

# The coefficient is a bit smaller, the standard error somewhat larger, so we
# introduced some bias, and reduced the power. Both could have been expected,
# due to how we generated missings and since we lost 373 cases.
```
### Regression Imputation
```{r}
imp_regression <- mice(
  data   = data_missing,
  method = "norm.predict",
  m      = 1,
  maxit  = 1
)
```

`complete()` takes an object, fills in the missing data, and returns the complete
data in a specified format.
```{r}
complete(imp_regression) |>  
  is.na() |>                
  sum()
```

```{r}
lm_regression <- with(imp_regression, lm(m_Expend ~ Outstate))

summary(lm_regression)
```

```{r}
imp_regression |> 
  ggmice(aes(x = Outstate, y = m_Expend)) +
  geom_point()

# Almost all imputed values are on a (almost) straight regression line (which 
# is how we specified the imputation model). The variation around this line
# comes from the other variables that are also used in calculating the
# predictions.
```

### Stochastic Regression Imputation

```{r}
imp_stochastic <- mice(
  data   = data_missing,
  method = "norm.nob",
  m      = 1,
  maxit  = 1
)
```

```{r}
lm_stochastic <- with(imp_stochastic, lm(m_Expend ~ Outstate))

summary(lm_stochastic)
```

```{r}
imp_stochastic |> 
  ggmice(aes(x = Outstate, y = m_Expend)) +
  geom_point()

# There is considerably more variation between the imputed values as compared to
# the regression imputations. However, the current approach still treats the 
# estimated regression coeffcients as fixed, although there is also uncertainty 
# with respect to these coefficients. This uncertainty is still not taken into 
# account with stochastic regression imputation.
```

### Multiple Imputation
```{r}
imp_multiple <- mice(data_missing) 

imp_multiple

```


```{r}
lm_multiple <- with(imp_multiple, lm(m_Expend ~ Outstate)) 

# The `lm_multiple` object contains the regression summaries for each of the 5
# datasets. The new object `lm_multiple` is actually of class `mira`, which 
# stands for Multiply Imputed Repeated Analyses.
# in the real data.
```


```{r}
pool_multiple <- pool(lm_multiple)  

summary(pool_multiple)

# These results show the regression coefficient and corresponding standard 
# error obtained after pooling the estimates from the five regression models
# applied on the five imputed datasets.

# The complete pooled object gives the relevant pooled regression coefficients 
# and parameters, as well as the fraction of information about the coefficients
# missing due to nonresponse (`fmi`) and the proportion of the variation
# attributable to the missing data (`lambda`). The pooled fit object is of 
# class `mipo`, which stands for multiply imputed pooled object.*
```

```{r}
imp_multiple |> 
  ggmice(aes(x = Outstate, y = m_Expend)) +
  geom_point()

# Here we take all uncertainty with respect to the imputations into account 
# (i.e., uncertainty with respect to the values and with respect to the
# coefficients). Moreover, we impute multiple times, so that we can assess the
# variability between different data sets (on the level of the estimated model
# parameters). This is often the best approach when interest is in making 
# inferences with respect to the imputed data.
```

## Comparisson
```{r}
data.frame("method" = c("complete data", 
                        "listwise deletion", 
                        "regression imputation",
                        "stochastic regression imputation",
                        "multiple imputation"),
           "coefficient" = c(summary(lm_complete)$coefficients[2,1],
                             summary(lm_listwise)$coefficients[2,1],
                             summary(lm_regression)$estimate[2],
                             summary(lm_stochastic)$estimate[2],
                             summary(pool_multiple)$estimate[2]),
           "standard error" = c(summary(lm_complete)$coefficients[2,2],
                                summary(lm_listwise)$coefficients[2,2],
                                summary(lm_regression)$std.error[2],
                                summary(lm_stochastic)$std.error[2],
                                summary(pool_multiple)$std.error[2]))

# When comparing the results from the different approaches, it can be 
# concluded that in case of a dataset where the missingness follows the 
# mechanism Seen Data Dependent, the estimate obtained after applying multiple
# imputation is closest to the original estimate obtained from the complete
# dataset. In addition, it can be concluded that the standard error obtained 
# using multiple imputation is somewhat larger than the standard error obtain 
# from the complete dataset. This is due to the fact that when the standard 
# errors are pooled over the different imputations, not only the average is 
# taken, but the variance over the differences is also incorporated (check 
# the pooling rules for more detail), i.e. the standard error also incorporates
# uncertainty due to the missing values. In contrast, the results obtained 
# using listwise deletion, regression imputation and stochastic imputation all 
# show a biased estimate, while regression imputation and stochastic 
# imputation also show a standard error that is too low.*

```


^